---
title: "Data Handling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Handling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```


**synthetic data will be added later and full example will be provided. **

# Description

This document gives idea what kind of data is correct to use with the functions and demonstrates few classifying functions.


```{r}
library(dplyr)
library(healthpopR)
```


## Load data

```{r}
# synthetic raw datasets
population <- arrow::read_parquet("../data/synth_population.parquet")
diagnoses <- arrow::read_parquet("../data/synth_diagnoses.parquet")
population_variables <- arrow::read_parquet("../data/synth_population_variables.parquet")
```


Population should have these variables:

| Variable Name | Description |
| :------------ | :--------------------------------------------------------------------- |
| ID | Patient identifier |
| DATE_BIRTH | Date of patient birth |
| DATE_DEATH | Date of when patient died or NA |
| DATE_MIGRATION | Date of when patient migrated to abroad or NA  |


Diagnoses data should have these variables:

| Variable Name | Description |
| :------------ | :--------------------------------------------------------------------- |
| ID | Patient identifier |
| DGREG | Diagnose Code Source (ICD10, ICD9 or ICD8) |
| SRC | Registry source (ex. hilmo, avohilmo etc.) |
| DATE | Date of the diagnose |
| DG | Full Diagnose Code (ex. E11, H903, H0410)|

But then we will create also these variables:

| ICD10_3LETTERS | **Only with ICD10 code.** Diagnose Code in first 3 letters |
| ICD10_CLASS | **Only with ICD10 code.** ICD-10 Diagnose Class (Ex. E00–E90, H60–H95, H00–H59) |
| DATE_BIRTH | Day of patient birth |
| AGE | **Only with murtumat** Patient age on diagnose date |


population_variables should have these variables:

| Variable Name | Description |
| :------------ | :--------------------------------------------------------------------- |
| ID | Patient identifier |
| edu | Education, in our example it factor with categories Low, Med and High |
| bmi | Body Mass Index as double |

For bmi we will categorized variables.


## Categorize ICD-10 Diagnoses

ICD-10 Codes are useful to categorize to 3 Letters and to upper category:

```{r}
diagnoses <- diagnoses %>% 
  mutate(
    ICD10_3LETTERS = healthpopR::recode_icd10_3letters(DG, DGREG = DGREG),
    ICD10_CLASS = healthpopR::recode_icd10_class(ICD10_3LETTERS, DGREG = DGREG)
  ) %>% 
  left_join(population %>% select(ID, DATE_BIRTH), by = "ID") %>% 
  mutate(AGE = floor(lubridate::time_length(difftime(DATE, DATE_BIRTH), "years")))
```



## Categorize population_variables BMI

BMI can be categorized to 4 or 6 category classes:

```{r}
population_variables <- population_variables %>% 
  mutate(
    bmi_cat1 = factor(healthpopR::categorize_bmi(bmi, levels = 6)),
    bmi_cat2 = factor(healthpopR::categorize_bmi(bmi, levels = 4)),
    edu = factor(edu)
  )

```

