---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

**This section is under development, because we don´t have a example data to show the results. Anyway, you can get idea of how functions are working.**

This example shows deeper detail of the `healthpopR` functions after your datasets are in right shape. Check section [Data Handling](articles/data-handling.html) first to get your data in right shape.

```{r setup}
library(healthpopR)
```


# Load Data 

Assuming that you have created datasets which are ready to use and analyzed. We can load datasets to use. 

```{r}
loc = "data/" 
population <- arrow::read_parquet(file = paste0(loc, "population.parquet"))
diagnoses <- arrow::read_parquet(file = paste0(loc, "diagnoses.parquet"))
data_dates <- arrow::read_parquet(file = paste0(loc, "data_dates.parquet"))
population_variables <- arrow::read_parquet(file = paste0(loc, "population_variables.parquet"))
```

Let's define some sample case of interest.

```{r}
selected_exposure_icd10 = "^E11"
selected_exposure_icd9 = "^250A"
selected_exposure_icd8 = "^250"
selected_exposure_regsrc = c("avohilmo", "erko", "hilmo", "local", "ksyy", "soshilmo", "syopa")
selected_response_icd10 = "^I2[0-5]"
selected_response_icd9 = "^41[0-4]"
selected_response_icd8 = "^41[0-4]"
selected_response_regsrc = c("avohilmo", "erko", "hilmo", "local", "ksyy", "soshilmo", "syopa")
```

# Diagnoses data 

Creating exposure diagnosis data 

```{r}
exposure_diagnoses <- search_diagnoses(
  regex_icd10 = selected_exposure_icd10,
  regex_icd9 = selected_exposure_icd9,
  regex_icd8 = selected_exposure_icd8,
  registry_source = selected_exposure_regsrc,
  data_diagnoses = diagnoses
)
```

Creating response diagnosis data

```{r}
response_diagnoses <- search_diagnoses(
  regex_icd10 = selected_response_icd10,
  regex_icd9 = selected_response_icd9,
  regex_icd8 = selected_response_icd8,
  registry_source = selected_response_regsrc,
  data_diagnoses = diagnoses
)
```

Checking what registry sources are found of the diagnoses

```{r}
plot_diagnoses_src(exposure_diagnoses)
plot_diagnoses_src(response_diagnoses, per_source = TRUE)
```



# Population Data 

Just exposure population without all the exposure codes 

```{r}
dpop_exposure <- classify_population(
  exposure_icd10 = selected_exposure_icd10,
  exposure_src = selected_exposure_regsrc)
```

Both, exposure and response population in one 

```{r}
dpop <- classify_population(exposure_icd10 = selected_exposure_icd10,
                     exposure_icd9 = selected_exposure_icd9,
                     exposure_icd8 = selected_exposure_icd8,
                     exposure_src = selected_exposure_regsrc,
                     response_icd10 = selected_response_icd10,
                     response_icd9 = selected_response_icd9,
                     response_icd8 = selected_response_icd8,
                     response_src = selected_response_regsrc)
```



# Population Analytics 

Exposure age distribution 

```{r}
plot_age_distribution(dpop, group = "exposure")
plot_age_distribution(dpop, group = "exposure", subgroups = TRUE)
```

Response age distribution 

```{r}
plot_age_distribution(dpop, group = "response")
plot_age_distribution(dpop, group = "response", subgroups = TRUE)
```

Age Distribution Tables 

```{r}
table_age_distribution(dpop, group = "exposure")
table_age_distribution(dpop, group = "response")
table_age_distribution(dpop, group = "response", subgroups = TRUE)
table_age_distribution(dpop, group = "exposure", subgroups = TRUE)
```

Summaries of Exposure Response Diagnoses 

```{r}
summary_exp_resp_order(dpop)
summary_exp_resp_crosstabulation(dpop, output = "viewer")
```

# Health Analytics

Creating A ICD-10 profile of the group

```{r}
health_profiles <- classify_icd10_profile(dpop, diagnoses = diagnoses, exposure_icd10 = selected_exposure_icd10, exposure_src = selected_exposure_regsrc)
plot_health_icd10_profile(health_profiles)
```

Finding top ICD-10 diagnoses on exposure group (compared to no-exposure group)

```{r}
tbl_health <- tbl_icd10_diff_by_exposure(dpop, diagnoses = diagnoses, exposure_icd10 = selected_exposure_icd10, exposure_src = selected_exposure_regsrc)
plot_icd10_diff_by_exposure(tbl_health)
```


# Survival Analysis

Creating a survival object and plotting Kaplan Meier and Competing Risk

```{r}
dsurv <- create_dsurv(dpop, 
                      censoring_date = as.Date("2023-12-21"), 
                      filter_early_responses = FALSE)
plot_survival_km(dsurv)
plot_survival_cr(dsurv)
```

# Cox Analysis

**This is under development and is currently relying on two extra datasets `data_dates` and `data_socioeconomic`, which are still under development of scaling to more universal format. **

Creating a cox dataset

```{r}
cox_base <- cox_create_data(dpop, 
                            data_dates = ostpre_vastpaiv, 
                            data_socioeconomic = edumiage,
                            reference_values = list("bmi_cat1" = "Healthy Weight", 
                                                    "bmi_cat2" = "Healthy Weight", 
                                                    "edu" = "3 - High"),
                            censoring_date = as.Date("2023-12-31"))
```

Plotting a survival element

```{r}
cox_plot_overall(cox_base, type = "event")
```


Creating a model for Cox analysis

```{r}
model1 <- create_cox_model(cox_base,  
                           normal_vars = c("edu"), 
                           spline_vars = c("age_bs", "bmi") 
                           )
```

Checking spline variable

```{r}
cox_plot_spline(model1, spline_var = "age_bs")
```

Other packages has some good functions to create plots

```{r}
ftest <- survival::cox.zph(model1)
survminer::ggcoxzph(ftest)
testmodel <- survival::coxph(as.formula(paste(deparse(model1$formula), collapse = " ")),
                   data =  cox_base)
survminer::ggforest(testmodel)
```

